Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-07-09 paper1_jiturra.Rmd
update

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
#   number: 1
# - name: Justin d'Ottawa
#   affiliation: University of Ottawa
#   number: 2
# - name: Pedro Torres
#   number: 1,2 
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez.  
abstract: | 
  | Escalating economic inequality has increased attention to the link between class relations and redistributive preferences in contemporary societies. However, economic inequality unevenly affects class relations and their influence on support for redistribution across societies. While those in the upper and middle classes reinforce their privileged access to resources and diverse networks, the working class has been disproportionately affected by social exclusion as inequality rises. Consequently, inequality consolidates segregated lifeworlds as it increases the divide between the lives and experiences of the different social classes.  Meeting people from diverse class positions can help individuals become more aware of other people's lifestyles and worldviews. Homogeneous upper-class networks may reduce empathy and solidarity towards those in need, resulting in decreased support for redistribution. Conversely, segregated lower-class social networks can lead to increased marginalization and reinforce support for redistribution. This study explores the relationship between class-based network segregation and redistributive preferences employing cross-national data from 32,529 individuals in 31 societies. The main findings suggest that social class conditions the association of network homogeneity on redistributive preferences, where working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity decreases support for redistribution. In addition, the conditional influence of network segregation weakens in unequal societies, especially for the upper classes. Implications for the study of class relations and political attitudes are discussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa6.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  # - \usepackage{times}           # Times New Roman
  - \usepackage{helvet}
  - \renewcommand{\familydefault}{\sfdefault} 
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

```{r eval=FALSE, include=FALSE}
 # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
 # clean #in the yml
 rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::pdf_document2")
 rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::html_document2")
 rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::word_document2")

@@

```{r include=FALSE}
wordcount<- rmdwc::rmdcount(files = "paper1_jiturra.Rmd")
wordcount$words
```

@@

During the past decades, the few cross-national studies on redistributive preferences and social class have predominantly focused on the situation of individuals and households [@lindh_class_2020]. Nonetheless, little attention has been paid to the network mechanisms involved in how individuals form their opinions regarding economic inequality and redistributive policies. On the one hand, sociologists have formerly demonstrated that homophily profoundly drives individuals to form similar social ties that consequently nurture social network segregation [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. In contrast, it is also known that opinions about economic inequality can be affected by class segregation as it provides a window for learning about the economic conditions, lifestyles, and worldviews of other social classes [@mijs_inequality_2018;@otero_power_2023]. Despite this, it can be argued that the scrutiny of redistributive preferences can be enhanced by broadening the comprehension of class relations from a network perspective, providing a comprehensive manner to understand processes of shared norm internalization, class identity formation, and collective solidarity [@Svallfors2006;@goldthorpe_sobre_1992]. In this regard, the present study examines the relationship between class-based network segregation and redistributive preferences in cross-national comparison, focusing on the role of economic inequality as a moderator of the claimed influence of class relations on support for redistribution.

While prior research has predominantly focused on examining the impact of social class through an individualistic lens, it is noteworthy that more attention should be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations that make them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. In addition, the normative basis of class relations introduces the relevance of the dimensions of solidarity and reciprocity, which have been argued to provide the moral basis for legitimacy and popular support for welfare schemes [@mau_moral_2003]. In this regard, besides the leverage of individual class position, recent efforts have pointed out that whereas the class positions of network ties do affect redistributive preferences, the degree of segregation in a homogeneous class environment can drive attitudes to become even stronger [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022].

While network segregation intensifies the already established class divide in redistributive preferences, little is known about how contextual factors affect the association of class-based network segregation and support for redistribution. On the one hand, studies on the class-attitude link posit that economic inequality is crucial for understanding how class relations are mirrored in redistributive demands as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding on this matter is that the upper classes tend to hold more egalitarian attitudes in contexts of high inequality, while the already high redistributive demands of the working class remain stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, income inequality leads to more segregated forms of social participation and network composition [@pichler_patterns_2007;@otero_space_2022]. Indeed, inequality reinforces the stratified access to social activities and widens the distance between classes, resulting in an increasing marginalization of the lower classes and consolidating the privileged positions of the upper classes that held better opportunities for diversifying their social activities and interpersonal ties [@lancee_income_2012;@letki_getting_2015;@otero_differences_2023]. Nevertheless, current efforts to examine the impact of income inequality have primarily concentrated on either social networks or support for redistribution. A key question that remains unanswered is the extent to which broader economic disparities can influence the relationship between class-based network segregation and support for redistribution within a comprehensive framework.

Thus, as no research has comparatively stressed the role of economic inequality in the relationship between class-based network segregation and redistributive preferences, the present study aims to answer two questions:

@@

For the present investigation, a data sample of 32,529 individuals across 31 societies from the International Social Survey Program (ISSP) 2017 is utilized. The ISSP presents a nonprecedence opportunity as it provides comprehensive information on interpersonal networks, social class, and attitudes toward redistribution.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020, p. 421]. In this view, the role of class goes beyond the market situation of individuals but drives collective economic interests and moral views about the role of the market and the state [@Svallfors2006]. Here, what has been understood as redistributive preferences entail the public responses to economic inequality regarding the sphere of the role of government actions in aspects such as the reduction of income differences or enhancing opportunities for those in economic despair  [@mccall_americans_2009].

However, class-based explanations of redistributive preferences have mainly been, but not exclusively, focused on the individual or household situation. It has been claimed that redistributive preferences can be explained through the labor market situation, which comprises access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009].
Moreover, as material interests might prevail in scarcity, value-driven motivations are a relevant explanatory factor of redistributive preferences [@Kulin2013;@feldman_humanitarian_2001], becoming salient under greater certainty and weaker under material precariousness [@Maldonadoetal2019]. Other approaches have suggested that given the substantial time workers spend performing their jobs, social relations in the workspace can imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. For instance, continuous and diverse social interactions that characterize interpersonal services can drive empathy and reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, the vertical monitoring in managerial occupations and the relevance of autonomy in the case of self-employed fortify self-interested and conservative political views [@langsaether_more_2020;@oesch_electoral_2018].

Empirically, the current understanding of the class divides on redistributive preferences is extensive [@andersen_preferences_2018;@lindh_public_2015; @curtis_how_2015;@langsaether_more_2020;@brooks_why_2010;@Svallfors2006]. Nevertheless, besides the market situation of individuals and their households, including class-based social ties raises the relevance of the _class situation_ as it comprises both economic interests and life chances that motivate communal action beyond the individualistic lens [@weber_class_2011, pp. 57-59]. Therefore, the role of social relations on redistributive preferences from a network perspective needs to be developed beyond mere resource interchanges, but as part of the sociability practices that structure class relations.

@@

It can be argued that class relations not only represent resource-based distinctions but also patterns of sociability observed in the differentiation of social ties. Class relations can also be understood as the degree of cross-class relationship as a structural characteristic of a society, representing a network of social ties between different class positions [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001, p.416]. Hereby, friendship and family ties are prone to be homogeneous in terms of class and demographic characteristics, but distant ties play the role of bridging individuals to other different social groups and contribute to network diversity [@bargsted_pautas_2020; @plaza_juntos_2022;@lazarsfeld_friendship_1954]. In this sense, socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004;@homans_human_1951]. Nonetheless, it can be argued that attitudinal similarity is the result of segregated social relations, which in turn are linked to structural processes of differentiation that also demarcate the boundaries to form ties with other social positions [@feld_focused_1981].

It is possible to find two approaches in the study of class relations from a network approach. On the one hand, it has emphasized the role of _diversity_ as the degree of connectedness to dissimilar occupations that vertically represent access to resources embedded in social networks [@lin_social_2007]. However, diversity is defined as the rate of dissimilar ties within the network that do not necessarily count with a reference position to describe the network composition. A second approach has underscored the role of _segregation_ as the lack of cross-class network ties. This perspective is conceptually closer to homophily as it is anchored in individual class positions and has empirically been addressed through the concept of network homogeneity [@otero_open_2021].

Empirically, studies on network diversity have shown that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the more homogeneous participation of the working class [@pichler_social_2009]. Similar patterns have been found in terms of the composition of social ties, where the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes [@cepic_how_2020;@carrascosa_class_2023]. Also, this stratified network structure holds during the life course, where the upper classes hold increasingly diverse social contacts in contrast to the stable networks of the working classes [@volker_social_2020]. In contrast, studies that have focused on segregation suggest that the property dimension is much less permeable than authority-based boundaries, arguing that class interests increase the social distance between proprietors and manual workers, while the intermediate class position of supervisors jointly to their higher contact frequency with manual workers make friendship tie formation more likely [@wright_relative_1992]. Similar studies have shown that the intermediate class holds higher permeability in contrast to the more homogeneous networks of the working class, suggesting that their limited life chances and lower capacity to be socially engaged ultimately result in a lack of social resources that lead to social segregation. In contrast, the upper class is less permeable and homogeneous because it tends to self-select as a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

Besides individual class-based mechanisms, it can be argued that their social relations represent a relevant factor in attitude formation. Particularly, despite limited research on the link between social networks and redistributive preferences, two theoretical approaches have discussed the role of social relations in attitude formation: reference groups and class-based networks.

It has been stated that perceptions about economic inequality rooted in social comparison processes with reference groups can explain the formation of redistributive preferences [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. The argument posits that people form their beliefs through family, friends, and coworkers' experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of these reference groups [@Evans1992, p. 467]. Therefore, inferences about the social world are linked to the degree of segregation in the immediate social environment of a person, which influences the intensity and character of the information that ultimately shapes inequality perceptions [@mijs_is_2021]. Accordingly, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more segregated networks [@summers_deliberating_2022]. Nevertheless, I argue that this body of research has been mainly focused on the cognitive dimension of preference formation through inequality perceptions or beliefs rather than straightforward addressing the claimed influence of network segregation on redistributive preferences [@cansunar_who_2021;@garcia-castro_perceived_2022;@Cruces2011;@becker_significant_2021].

In contrast, a network approach provides a better picture of class relations that nurture social norms and group identity [@kalmijn_social_2007, p. 550]. More precisely, it has been argued that redistributive preferences are influenced by the class situation of both the individual and network ties, where social influence processes can either allineate or divide opinions according to the contacts' class position and how segregated contact opportunities are [@vargassalfate_contact_2023]. This claim resembles the fact that classes are characterized as collectivities with differences in their degrees of cohesion and solidarity, encompassing unequal status-based social interactions that are linked to individual and household material well-being, cultural perspectives, and political preferences that structure broader social experiences [@morris_attenuation_1996, p. 48]. Furthermore, social integration can be affected in societies with lower contact opportunities between different social classes, creating an "empathy gulf" that comprises barriers to imagining others' lifestyles in contexts of rising inequality [@sachweh_moral_2012]. Thus, spatially segregated interactions may nurture doubts about worthiness when the lower classes contrast themselves with the lifestyles of the upper classes, which in turn can undermine feelings of social inclusion and cohesion [@sachweh_moral_2012]. As a result, segregation drives the lives of others to become more distant and might have consequences for empathy and solidarity toward others, potentially leading to the perception of fellow citizens as strangers [@otero_lives_2022, p. 758].

@@

The class position of surrounding family members, friends, and acquaintances influences redistributive preferences as they function as socialization agents that can be bolstered in segregated social networks. In principle, political attitudes are connected to class interests and norms as they are nurtured in the family of origin during childhood and early adulthood. For instance, it has been shown that individuals with network ties to the upper class through parental connections support redistribution and progressive taxation less than those from working-class family backgrounds [@lee_consider_2023]. Also, as households share risk according to the class position of their members, redistributive preferences are not only affected by their family background but also by partners' class positions. For example, @paskov_crossclass_2022 found that working-class ties bolster redistributive preferences but decrease with upper-class ties, becoming much more intense when the class positions of individuals, partners, and parents form a more homogeneous network. Besides, @lindh_missing_2021 found that including friends and acquaintances in the network structure leads individuals with higher ties to the managerial class to decrease their redistributive preferences compared to those with higher sociocultural and working-class ties, suggesting that people tend to consider and adjust their attitudes according to the class position of their contacts [@lindh_missing_2021, p. 698].

In summary, the tendency to form homogenous social ties regarding social class is expected to reinforce attitudes and consolidate opinion similarity. Empirically, a weak direct association between homogeneity and redistributive preferences is expected because class segregation does not distinguish between classes as it provides only the _overall_ degree of segregation. In contrast, the focus is on how class-based network homogeneity is conditional to social class. Empirically, this can be observed through the interaction between network homogeneity and social class. Precisely, this will indicate if being segregated into lower (upper) classes leads to stronger (weaker) redistributive preferences ($H_1$).

@@

Studies that have stressed the role of economic inequality on the relationship between social class and redistributive preferences indicate that the lower and the upper classes react differently to rising economic inequality. Theoretically, political economists have suggested that high-income individuals are far from monolithic in their redistributive preferences, arguing that their concerns about the harmful _consequences_ of economic inequality (e.g., crime) ultimately motivate altruistic support for redistribution [@rueda_food_2018;@dimick_models_2018;@dimick_altruistic_2017;@rueda_externalities_2016]. Conversely, the moral economy literature in sociology has argued that the differences among the affluent can be explained as a matter of distributive justice evaluations about the _procedures_ for resource allocation [@liebig_sociology_2016;@atria_economic_2020]. Hence, affluent groups are more responsive because they perceive that increasing inequality harms the overall opportunity structure and social mobility chances [@sachweh_criticizing_2017]. Likewise, lower perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of justice in the conditions for getting ahead [@kim_socioeconomic_2018]. In contrast, low-income individuals perceive ascribed characteristics as more important in constraining the opportunity structure, regardless of current income inequality [@sachweh_why_2019, p. 656].

Regarding social networks, prior studies have argued that social relations are directly linked to resource access and how these are distributed in society. Here,  Neckerman & Torche [-@neckerman_inequality_2007, p. 344] suggest that experiencing marginalization is deeply associated with the life chances to participate openly in social life, which can be exacerbated in contexts of greater material inequality. Also, economic inequality can lead to greater perceived status competition [@wilkinson_spirit_2010;@garcia-sanchez_perceived_2024], which in turn hampers trustworthy social ties among the low-status individuals [@salgado_expectations_2021]. In contrast, societies with robust welfare institutions and lower economic inequality foster social engagement in civic and social activities, thereby enhancing social trust between citizens and bolstering social solidarity [@uslaner_inequality_2005;@kragten_income_2017].

>>>>

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado  
#linestretch: '2.0'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
abstract: | 
  | Rising economic inequality has increased attention to the link between class relations and redistributive preferences in contemporary societies. However, economic inequality unevenly affects class relations and their influence on support for redistribution across societies. While those in the upper and middle classes reinforce their privileged access to resources and diverse networks, the working class has been disproportionately affected by social exclusion as inequality rises. Consequently, inequality consolidates segregated lifeworlds as it increases the divide between the lives and experiences of the different social classes.  Meeting people from diverse class positions can help individuals become more aware of other people's lifestyles and worldviews. Homogeneous upper-class networks may reduce empathy and solidarity towards those in need, resulting in decreased support for redistribution. Conversely, segregated lower-class social networks can lead to increased marginalization and reinforce support for redistribution. This study explores the relationship between class-based network segregation and redistributive preferences employing cross-national data from 32,529 individuals in 31 societies. The main findings suggest that social class conditions the association of network homogeneity on redistributive preferences, where working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity decreases support for redistribution. In addition, the conditional influence of network segregation weakens in unequal societies, especially for the upper classes. Implications for the study of class relations and political attitudes are discussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa-no-doi-no-issue.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  # - \usepackage{times}           # Times New Roman
  - \usepackage{helvet}
  - \usepackage{endnotes}
  - \let\footnote=\endnote 
  - \renewcommand{\familydefault}{\sfdefault} 
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

```{r eval=FALSE, include=FALSE}
 # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
 # clean #in the yml
 rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::pdf_document2")
 # rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::html_document2")
 rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::word_document2")

@@

```{r include=FALSE, cache=FALSE}
wordcount<- rmdwc::rmdcount(files = "paper1_jiturra.Rmd") 
wordcount$words
```

@@

Over the past decades, cross-national studies on redistributive preferences and social class have predominantly focused on individuals and households [@lindh_class_2020]. In contrast, less attention has been paid to how social networks influence redistributive preferences. Sociologists have demonstrated that individuals tend to segregate social networks [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Additionally, opinions about economic inequality can be affected by class segregation, as it provides insight into the economic conditions, lifestyles, and worldviews of other social classes [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. Despite this, the study of redistributive preferences can be enhanced by broadening the understanding of class relations from a network perspective. This approach offers a comprehensive understanding of processes such as shared norm internalization, class identity formation, and collective solidarity [@Svallfors2006;@goldthorpe_sobre_1992]. Thus, this study examines the relationship between class-based network segregation and redistributive preferences in a cross-national comparison, focusing on the role of economic inequality as a moderator in the influence of class relations on support for redistribution.

While prior research has predominantly examined the impact of social class through an individualistic lens, more attention should be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations, making them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. Furthermore, the normative basis of class relations introduces the relevance of solidarity and reciprocity dimensions, which are argued to provide the moral foundation for legitimacy and popular support for welfare schemes [@mau_moral_2003]. In this regard, beyond the influence of individual class position, recent efforts have highlighted that while the class positions of network ties do affect redistributive preferences, the degree of segregation in a homogeneous class environment can drive attitudes to become even stronger [@otero_lives_2022].

While network segregation widens the class divide in redistributive preferences, little is known about how income inequality can affect the association between class-based network segregation and support for redistribution. On the one hand, studies on the class-attitude link suggest that income inequality is crucial for understanding how class relations are reflected in redistributive demands, as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding in this literature is that the upper classes tend to support redistribution more in contexts of high inequality, while the already high redistributive demands of the working class remain stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, income inequality leads to more segregated forms of social participation and network composition [@pichler_patterns_2007;@otero_space_2022]. Indeed, inequality reinforces stratified access to social activities and widens the distance between classes, resulting in increased marginalization of the lower classes and consolidating the privileged positions of the upper classes, who have better opportunities to diversify their social activities and interpersonal ties [@lancee_income_2012;@otero_differences_2023]. Nevertheless, current efforts to examine the impact of income inequality have primarily focused on either social networks or support for redistribution. A key question that remains unanswered is the extent to which broader economic disparities moderate the relationship between class-based network segregation and support for redistribution.

Thus, to the best of my knowledge, no research has comparatively investigated the role of economic inequality in the relationship between class-based network segregation and redistributive preferences. This study aims to address two key questions:

@@

For this investigation, I utilized a sample of 32,529 individuals across 31 countries from the International Social Survey Program (ISSP) 2017. This dataset provides unprecedented cross-national comparative data on social networks, social class, and attitudes toward redistribution.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020, p. 421]. In this view, social class not only represents individuals’ market situations but also their economic interests and moral perspectives on the role of the market and the state in the distribution and redistribution of resources [@Svallfors2006]. Here, redistributive preferences are understood as the support for policies and mechanisms aimed at reducing economic inequality through the redistribution of wealth and resources [@mccall_americans_2009]. These preferences encompass views on taxation, welfare programs, public services, and other government interventions designed to transfer resources from wealthier to less wealthy individuals or groups in society [@garcia-sanchez_two_2022].

Class-based explanations of redistributive preferences have primarily focused on the individual or household level but are not exclusively limited to these contexts. It has been argued that redistributive preferences can be explained through labor market situations, which encompass access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009]. Additionally, while material interests may dominate in situations of scarcity, value-driven motivations become a significant explanatory factor for redistributive preferences under conditions of greater certainty and are weaker under material precariousness [@Kulin2013;@Maldonadoetal2019]. Other approaches suggest that the substantial time workers spend on their jobs allows social relations in the workplace to imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. For example, the continuous and diverse social interactions inherent in interpersonal services can foster empathy and reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, vertical monitoring in managerial occupations and the emphasis on autonomy in self-employed roles tend to strengthen self-interested and conservative political views [@oesch_electoral_2018].

Empirically, the current understanding of class divides in redistributive preferences is extensive [@lindh_public_2015;@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010]. Nevertheless, beyond the market situation of individuals and their households, considering class-based network ties highlights the relevance of the class situation, as it encompasses both economic interests and life chances that drive communal action beyond individuals [@weber_class_2011, pp. 57-59]. Thus, the role of social relations in shaping redistributive preferences from a network perspective considers more straightforwardly the social ties that structure class relations.

@@

<!-- It can be argued that class relations not only represent resource-based distinctions but also patterns of connectedness observed in the differentiation of social ties.  -->

Class relations can also be understood as the degree of cross-class relationships as a structural characteristic of society, representing a network of social ties between different class positions [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001, p.416]. In this context, friendship and family ties tend to be homogeneous concerning class and demographic characteristics, while more distant ties serve to bridge individuals to different social groups and contribute to network diversity [@diprete_segregation_2011; @lazarsfeld_friendship_1954]. Besides, socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004; @homans_human_1951]. Nevertheless, attitude similarity in segregated networks is rather a consequence of the social differentiation that structures the formation of cross-class network ties [@feld_focused_1981].

There are two primary approaches in the study of class relations from a network perspective. On the one hand, network diversity is defined as the degree of connectedness to dissimilar occupations, which represents vertical access to resources embedded in social networks [@lin_access_1986]. However, diversity is defined as the rate of dissimilar ties within the network without necessarily providing a reference position to describe network similarity. On the other hand, network segregation is the absence of cross-class connections and represents how similar network ties are in contrast to the individual. This perspective is conceptually closer to homophily as it is anchored in individual class positions and has been empirically addressed through network homogeneity [@otero_open_2021].

How does social class stratify social networks? Empirically, studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper-class forming connections with diverse individuals, in contrast to the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are observed in the composition of social ties, where the upper and intermediate classes maintain increasingly diverse and prestigious social environments compared to the working class [@cepic_how_2020;@carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest since class interests widen the social distance between proprietors and manual workers, the intermediate position of supervisors, along with their higher contact frequency with manual workers, makes the formation of friendship ties more likely. Similarly, @otero_open_2021 have shown that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class, suggesting that their limited life chances and lower capacity for social engagement ultimately result in a lack of social resources that leads to segregation. Conversely, the upper class is less permeable and homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

Besides individualistic class-based mechanisms in redistributive preferences, the argument that network ties have implications for attitude formation is not novel. In particular, two approaches have discussed the role of social relations on redistributive preferences: reference groups and class-based networks.

It has been stated that perceptions about economic inequality rooted in social comparison processes with reference groups can explain the formation of redistributive preferences [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. The argument posits that people form their beliefs through family, friends, and coworkers' experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of these reference groups [@Evans1992, p. 467]. Therefore, inferences about the social world are linked to the degree of segregation in the immediate social environment of a person, which influences the intensity and character of the information that ultimately shapes inequality perceptions [@mijs_is_2021]. Accordingly, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more segregated networks [@summers_deliberating_2022]. Nevertheless, I argue that this body of research has been mainly focused on the cognitive dimension of preference formation through inequality perceptions rather than straightforwardly addressing the claimed influence of network segregation on redistributive preferences [@cansunar_who_2021;@garcia-castro_perceived_2022].

In contrast, social networks provide a comprehensive picture of the class relations that contribute to group identity formation and internalization of social norms [@kalmijn_social_2007, p. 550]. Specifically, it has been argued that redistributive preferences are influenced not only by individuals' class situations but also by the class positions of their network ties [@paskov_crossclass_2022]. Social influence processes can either align or divide opinions depending on the class positions of contacts and the level of network segregation [@lindh_missing_2021]. These findings reflect the notion that classes are characterized as collectivities with varying degrees of cohesion and solidarity, comprising asymmetric status-based interactions related to material resources, cultural practices, and political preferences that shape broader social experiences [@morris_attenuation_1996, p. 48]. Following @sachweh_moral_2012, social integration can be impeded in societies with few opportunities for contact between different social classes, creating an "empathy gulf" that hinders individuals from understanding others' lifestyles amid rising inequality. Consequently, spatially segregated interactions may lead individuals to view the lives of different classes as more distant, which can undermine feelings of social inclusion and cohesion [@sachweh_moral_2012]. Thus, segregation can exacerbate perceptions of others as strangers and may impact empathy and solidarity, potentially diminishing social cohesion [@otero_lives_2022, p. 758].

@@

The class position of surrounding family members, friends, and acquaintances influences redistributive preferences, as these individuals function as socialization agents whose impact can be amplified in segregated social networks. In principle, political attitudes are connected to class interests and norms that are nurtured in the family of origin during childhood and early adulthood. For instance, @lee_consider_2023 shows that individuals with network ties to the upper class through parental connections tend to support redistribution and progressive taxation less than those from working-class family backgrounds. Moreover, since households share risk based on the class position of their members, redistributive preferences are shaped not only by family background but also by the class positions of partners. For example, @paskov_crossclass_2022 indicates that working-class ties bolster redistributive preferences, whereas ties with the upper class decrease these preferences, with the effects becoming more pronounced when the class positions of individuals, partners, and parents create a more homogeneous network. Additionally, @lindh_missing_2021 found that friendship and acquaintanceship ties to the managerial class are associated with lower redistributive preferences compared to ties with the sociocultural and working classes. Hence, they suggest that individuals tend to adjust their attitudes based on the class position of their contacts [@lindh_missing_2021, p. 698].

In summary, as homogeneity refers to the overall degree of segregation without distinguishing between class positions I anticipate a weak direct association between network homogeneity and redistributive preferences. Conversely, I hypothesize that the association of network homogeneity with redistributive preferences is conditional on social class as homogeneous social networks should reinforce attitude similarity (*segregation hypothesis*). Specifically, I propose that greater network segregation in the lower classes is associated with higher redistributive preferences, whereas greater segregation in the upper classes is related to lower redistributive preferences. Therefore, the first hypothesis is as follows:

> H1: _The greater the network segregation in the lower (upper) classes, the higher (lower) the redistributive preferences. _

@@

Studies on economic inequality, social class, and redistributive preferences indicate that the social classes react differently to rising economic inequality. Theoretically, political economists have suggested that high-income individuals are far from monolithic in their redistributive preferences, arguing that their concerns about the harmful _consequences_ of economic inequality (e.g., crime) ultimately motivate altruistic support for redistribution [@dimick_models_2018;@dimick_altruistic_2017;@rueda_externalities_2016]. Conversely, the moral economy literature in sociology has argued that the differences among the affluent can be explained as a matter of distributive justice evaluations about the _procedures_ for resource allocation [@liebig_sociology_2016].

Empirically, affluent groups are more responsive because they perceive the overall opportunity structure and social mobility chances are affected by rising inequality [@sachweh_why_2019]. Likewise, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of justice in the conditions for getting ahead [@kim_socioeconomic_2018]. In contrast, low-income individuals perceive ascribed characteristics as more important in constraining the opportunity structure, regardless of current income inequality [@sachweh_why_2019]. In a similar perspective, @edlund_democratic_2015 argue that in unequal societies with residual welfare states social class has a poor political meaning, making distributive conflicts weakly institutionalized around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can also be explained as they are often more aware of the institutional incentives and societal consequences of income inequality [@curtis_how_2015;@Svallfors2006].
